---
title: "Week 6 Discussion: Solution"
subtitle: "Saving plots"
description: "Tuesday February 11^th^, 2025"
---

## Setup 

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(showtext)

#......................import Google fonts.......................
# `name` is the name of the font as it appears in Google Fonts
# `family` is the user-specified id that you'll use to apply a font in your ggpplot
font_add_google(name = "Josefin Sans", family = "josefin")
font_add_google(name = "Sen", family = "sen")

#....................import Font Awesome fonts...................
font_add(family = "fa-brands",
         regular = here::here("fonts", "ARCHIVE", "Font Awesome 6 Brands-Regular-400.otf"))
font_add(family = "fa-regular",
         regular = here::here("fonts", "ARCHIVE", "Font Awesome 6 Free-Regular-400.otf")) 
font_add(family = "fa-solid",
         regular = here::here("fonts", "ARCHIVE", "Font Awesome 6 Free-Solid-900.otf"))

#................enable {showtext} for rendering.................
showtext_auto()
showtext_opts(dpi = 300) #<1>

#..........................import data...........................
jobs <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")
```
1. You'll want `{showtext}`'s rendering parameters to match those of `ggsave()`. Set both to a DPI of 300 (often used as a standard for quality print images).

## Wrangle data

```{r}
#| eval: true
#| echo: true
#| warning: false
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                wrangle data                              ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

jobs_clean <- jobs |>

 # add col with % men in a given occupation (% females in a given occupation was already included) ----
  mutate(percent_male = 100 - percent_female) |>
         
  # rearrange columns ----
  relocate(year, major_category, minor_category, occupation,
          total_workers, workers_male, workers_female,
          percent_male, percent_female,
          total_earnings, total_earnings_male, total_earnings_female, 
          wage_percent_of_male) |> 

  # drop rows with missing earning data ----
  drop_na(total_earnings_male, total_earnings_female) |>

  # make occupation a factor (necessary for reordering groups in our plot) ----
  mutate(occupation = as.factor(occupation)) |>

  # classify jobs by percentage male or female (these will become facet labels in our plot) ----
  mutate(group_label = case_when(
    percent_female >= 75 ~ "Occupations that are 75%+ female",
    percent_female >= 45 & percent_female <= 55 ~ "Occupations that are 45-55% female",
    percent_male >= 75 ~ "Occupations that are 75%+ male"
  ))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                              create subset df                            ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#....guarantee the same random samples each time we run code.....
set.seed(0)

#...............get random samples from each group...............

# 10 random jobs that are 75%+ female (2016) ----
f75 <- jobs_clean |>
  filter(year == 2016, group_label == "Occupations that are 75%+ female") |>
  slice_sample(n = 10)

# 10 random jobs that are 75%+ male (2016) ----
m75 <- jobs_clean |>
  filter(year == 2016, group_label == "Occupations that are 75%+ male") |>
  slice_sample(n = 10)

# 10 random jobs that are 45-55%+ female (2016) ----
f50 <- jobs_clean |>
  filter(year == 2016, group_label == "Occupations that are 45-55% female") |>
  slice_sample(n = 10)

#.......combine dfs & relevel factors (for plotting order).......
subset_jobs <- rbind(f75, m75, f50) |>
  mutate(group_label = fct_relevel(group_label, 
                                   "Occupations that are 75%+ female",
                                   "Occupations that are 45-55% female", 
                                   "Occupations that are 75%+ male"))
```

## Recreate original plot 

```{r}
#| eval: true
#| echo: true
#| warning: false
# original plot ----
plot <- ggplot(subset_jobs) +
  geom_segment(aes(x = total_earnings_female, 
                   xend = total_earnings_male,
                   y = fct_reorder(occupation, total_earnings), 
                   yend = occupation)) +
  geom_point(aes(x = total_earnings_male, y = occupation),
             color = "#CD93D8", 
             size = 3.25) +
  geom_point(aes(x = total_earnings_female, y = occupation),
             color = "#6A1E99", 
             size = 3.25) +
  facet_wrap(~group_label, nrow = 3, scales = "free_y") +
  scale_x_continuous(labels = scales::label_dollar(scale = 0.001, suffix = "k"),
                     breaks = c(25000, 50000, 75000, 100000, 125000))

plot
```

## Create new palette

```{r}
#| eval: true
#| echo: true
#| warning: false
# create a named color palette ----
earnings_pal <- c("males" = "#2D7787",
                  "females" = "#FC6B4B",
                  "dark_text" = "#0C1509",
                  "light_text" = "#4E514D") 

# preview it -----
monochromeR::view_palette(earnings_pal)
```

## Update plot colors

```{r}
#| eval: true
#| echo: true
#| warning: false
# apply updated colors ----
plot <- ggplot(subset_jobs) +
  geom_segment(aes(x = total_earnings_female, xend = total_earnings_male,
                   y = fct_reorder(occupation, total_earnings), yend = occupation)) +
  geom_point(aes(x = total_earnings_male, y = occupation),
             color = earnings_pal["males"], size = 3.25) +
  geom_point(aes(x = total_earnings_female, y = occupation),
             color = earnings_pal["females"], size = 3.25) +
  facet_wrap(~group_label, nrow = 3, scales = "free_y") +
  scale_x_continuous(labels = scales::label_dollar(scale = 0.001, suffix = "k"),
                     breaks = c(25000, 50000, 75000, 100000, 125000))

plot
```

## Add titles & theme

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
#| fig-asp: 1
#| out-width: "100%"
#.........................create caption.........................
github_icon <- "&#xf09b"
github_username <- "yourGitHubUserName"

caption <- glue::glue(
  "Data Source: TidyTuesday (March 5, 2019)<br>
  <span style='font-family:fa-brands;'>{github_icon};</span>
  {github_username}"
)

#........................create subtitle.........................
money_icon <- "&#xf3d1"

subtitle <- glue::glue("Median earnings 
                       <span style='font-family:fa-regular;'>{money_icon};</span>
                       of full-time
                       <span style='color:#2D7787;'>**male**</span> 
                       versus 
                       <span style='color:#FC6B4B;'>**female**</span>
                       workers by occupation in 2016")

#..........................modify plot...........................
updated_plot <- plot +
  labs(title = "Males earn more than females across most occupations",
       subtitle = subtitle,
       caption = caption) +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(family = "josefin",
                              face = "bold",
                              size = 18),
    plot.subtitle = ggtext::element_textbox(family = "sen",
                                            size = 11.5,
                                            color = earnings_pal["light_text"],
                                            #fill = "lightblue", # for debugging #<1>
                                            width = unit(25, "cm"),
                                            margin = margin(t = 2, r = 0, b = 6, l = 0),
                                            padding = margin(t = 5, r = 0, b = 5, l = 0)),
    plot.caption = ggtext::element_textbox(family = "sen",
                                           face = "italic", # NOTE: this no longer applies since the typeface "sen" does not exist in an italic font style
                                           color = earnings_pal["light_text"],
                                           #fill = "lightblue", # for debugging 
                                           halign = 1, 
                                           lineheight = 1.5,
                                           width = unit(25, "cm"),
                                           margin = margin(t = 15, r = 0, b = 0, l = 0),
                                           padding = margin(t = 10, r = 0, b = 10, l = 0)),
    strip.text.x = element_text(family = "josefin",
                                face = "bold",
                                size = 10,
                                hjust = 0),
    panel.spacing.y = unit(0.5, "cm"),
    axis.text = element_text(family = "sen",
                             size = 9,
                             color = earnings_pal["light_text"]),
    axis.title = element_blank()
  )

updated_plot
```
1. Adding fill color to `ggtext::element_textbox()`s can help to identify if the size is as intended once saved using `ggsave()` 

::: {.callout-note}
## Reminder: build your viz for your desired output!
Notice that making these modifications means we get a plot that doesn't render as nicely in our Quarto doc (e.g. y-axis text feels a bit more crowded, some of the panel strip text gets cut off). This is a good reminder that your final desired output format will influence how you build / modify your plot!
:::

## Save plots using `ggsave()`

### 8"x7" plot

```{r}
#| eval: false
#| echo: true
# save 8"x7" png ---
ggsave(
  filename = here::here("week6", "images", "salary-plot-8x7.png"),
  plot = updated_plot,
  device = "png",
  width = 8,
  height = 7,
  unit = "in",
  dpi = 300, #<1>
  create.dir = TRUE #<2> 
)
```
1. Default DPI is 300 (so you don't *need* to explicitly include this argument)
2. Creates a new directory at specified file path 

### 14"x12" plot

```{r}
#| eval: false
#| echo: true
# save 14"x12" png ---
ggsave(
  filename = here::here("week6", "images", "salary-plot-14x12.png"),
  plot = updated_plot,
  device = "png",
  width = 14, 
  height = 12,
  unit = "in",
  dpi = 300,
  scale = 0.6
)
```

## Turn off showtext rendering

```{r}
# turn off showtext text rendering ----
showtext_auto(FALSE)
```
