---
title: "KEY - interactive data visualization"
subtitle: "plotly & htmlwidgets"
author: "Your Name"
date: 2026-03-4
toc: true
---

::: {.callout-note}
This template follows the [interactive data visualization](../lecture-slides/SLIDES-interactivity.qmd){target="_blank"} slides. Please be sure to cross-reference the slides, which contain important conceptual context and comparisons!
:::

## Setup

::: {.callout-note}
We use the `storms` dataset from `{dplyr}` throughout this lab — no external file needed. It contains Atlantic storm tracks with wind speed, pressure, storm status, and geographic position from 1975 to 2022.
:::

<!-- CODE RENDERS BUT DOES NOT RUN -->

```{r setup-echo}
#| eval: false
#| echo: true
#| warning: false
#| message: false

#..........................load packages.........................
library(tidyverse)
library(plotly)
library(htmlwidgets)
library(scales)

#.......................preview the dataset.......................
glimpse(storms)
```

<!-- CODE RUNS BUT DOES NOT RENDER (FOR FILE PATH) -->

```{r setup-eval}
#| eval: true
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(plotly)
library(htmlwidgets)
library(scales)
```

**Key variables in `storms`:**

| Variable | Type | Description |
|---|---|---|
| `name` | character | Storm name |
| `year`, `month`, `day`, `hour` | numeric | Date & time of observation |
| `lat`, `long` | numeric | Storm center position |
| `status` | factor | `"tropical depression"`, `"tropical storm"`, `"hurricane"`, etc. |
| `category` | numeric | Saffir-Simpson category (1–5; `NA` for non-hurricanes) |
| `wind` | numeric | Max sustained wind speed (knots) |
| `pressure` | numeric | Min central pressure (millibars) |
| `tropicalstorm_force_diameter` | numeric | Diameter (nautical miles) of TS-force winds |

```{r preview-data}
#| eval: true
#| echo: true
#| message: false

glimpse(storms)
```

## Wrangle data

```{r}
storms_clean <- storms %>%
  mutate(status = tools::toTitleCase(gsub("-", " ", status))) %>%
  filter(!is.na(wind), !is.na(pressure), !is.na(status)) %>%
  filter(wind > 0, pressure > 800, pressure < 1030) %>% 
    group_by(name, year, status) %>%
  summarise(
    wind     = mean(wind, na.rm = TRUE),
    pressure = mean(pressure, na.rm = TRUE),
    .groups  = "drop"
  )

```



## Part 1: `ggplotly()` — converting a static ggplot

See slides: [Two approaches](../lecture-slides/SLIDES-interactivity.qmd#two-approaches){target="_blank"} and [ggplotly()](../lecture-slides/SLIDES-interactivity.qmd#ggplotly-how){target="_blank"}

### Build a static ggplot

We start with a jitter plot of wind speed vs. pressure.


```{r}
#| eval: true
#| echo: true
#| warning: false
#| fig-align: "center"
#| out-width: "80%"


ggplot_static <- storms_clean %>%
  ggplot(aes(
    x     = pressure,
    y     = wind,
    color = status,
    text  = paste0(
      "<b>", name, " (", year, ")</b><br>",
      "Status: ", status, "<br>",
      "Wind: ", round(wind,1), " kt<br>",
      "Pressure: ", round(pressure,1), " mb"
    )
  )) +
  geom_jitter(alpha = 0.35, size = 0.9, width = 1.5, height = 1.5) +
  scale_color_brewer(palette = "Set3") +
  labs(
    title  = "Atlantic Storm Intensity by Status",
    x      = "Pressure (mb)",
    y      = "Wind Speed (knots)",
    color  = "Status"
  ) +
  theme_minimal(base_size = 13)

ggplot_static
```



### Convert to interactive with `ggplotly()`

```{r}
ggplotly(ggplot_static, tooltip = "text") 
```

::: {.callout-tip}
**Try it:** 

- Hover over any point to see the formatted label.
- Click a status category in the legend to hide it.
- Double click a status category to view only that category. 
- Drag to zoom; double-click to reset. 

The toolbar in the top-right gives additional controls!
:::

### Customising `ggplotly()` output

A few small tweaks make `ggplotly()` output cleaner:

```{r ggplotly-polished}
#| eval: true
#| echo: true
#| warning: false

ggplotly(ggplot_static, tooltip = "text") |>
  layout(
    hoverlabel = list(bgcolor = "white", bordercolor = "grey80",
                      font = list(size = 12))
  ) |>
  config(displaylogo = FALSE,
         modeBarButtonsToRemove = list("select2d", "lasso2d"))
```

---

## Part 2: `plot_ly()` — native plotly for more control

See slides: [plot_ly() grammar](../lecture-slides/SLIDES-interactivity.qmd#plot-ly-grammar){target="_blank"} and [animation](../lecture-slides/SLIDES-interactivity.qmd#animation-intro){target="_blank"}

### Bar chart: storm counts by status and decade

```{r bar-chart}
#| eval: true
#| echo: true
#| warning: false

storms_clean |>
  mutate(decade = paste0(floor(year / 10) * 10, "s")) |>
  count(decade, status) |>
  plot_ly(
    x             = ~decade,
    y             = ~n,
    color         = ~status,
    colors        = RColorBrewer::brewer.pal(8, "Paired"),
    type          = "bar",
    text          = ~paste0("<b>", status, "</b><br>n = ", n),
    textposition  = "none",
    hoverinfo     = "text",
    hovertemplate = "%{text}<extra></extra>"
  ) |>
  layout(
    barmode = "stack",
    title   = "Atlantic Storm Counts by Decade and Status",
    xaxis   = list(title = "Decade"),
    yaxis = list(title = "", showticklabels = FALSE),
    legend  = list(title = list(text = "<b>Status</b>"))
  )
```
```{r}
#| warning: false
#| message: false

plot_ly(storms, lat = ~lat, lon = ~long,
        frame = ~year, type = "scattergeo", mode = "markers",
        color = ~as.factor(category), colors = RColorBrewer::brewer.pal(5, "BuPu"),
        marker = list(size = 6, opacity = 0.7),
        text = ~paste0("<b>", name, "</b><br>Category: ", category),
        hoverinfo = "text") |>
  layout(
    title = "Atlantic Storm Positions by Year",
    geo = list(
      scope        = "north america",
      showland     = TRUE,  landcolor  = "#2d3436",
      showocean    = TRUE,  oceancolor = "#1a1a2e",
      showcoastlines = TRUE, coastlinecolor = "#636e72",
      showlakes    = TRUE,  lakecolor  = "#1a1a2e",
      bgcolor      = "#1a1a2e",
      projection   = list(type = "mercator"),
      lonaxis      = list(range = c(-110, -10)),
      lataxis      = list(range = c(5, 60))
    ),
    paper_bgcolor = "#1a1a2e",
    font = list(color = "white")
  ) |>
  animation_opts(frame = 1000, transition = 0)
```

---

## Part 3: Saving and embedding widgets

See slides: [saveWidget()](../lecture-slides/SLIDES-interactivity.qmd#htmlwidgets-save){target="_blank"} and [embedding in Quarto](../lecture-slides/SLIDES-interactivity.qmd#quarto-embed){target="_blank"}

### Save a widget to a standalone HTML file

```{r save-widget}
#| eval: false
#| echo: true

#...............assign the animated chart to an object...........
animated_storms <- plot_ly(storms, lat = ~lat, lon = ~long,
        frame = ~year, type = "scattergeo", mode = "markers",
        color = ~as.factor(category), colors = RColorBrewer::brewer.pal(5, "BuPu"),
        marker = list(size = 6, opacity = 0.7),
        text = ~paste0("<b>", name, "</b><br>Category: ", category),
        hoverinfo = "text") |>
  layout(
    title = "Atlantic Storm Positions by Year",
    geo = list(
      scope        = "north america",
      showland     = TRUE,  landcolor  = "#2d3436",
      showocean    = TRUE,  oceancolor = "#1a1a2e",
      showcoastlines = TRUE, coastlinecolor = "#636e72",
      showlakes    = TRUE,  lakecolor  = "#1a1a2e",
      bgcolor      = "#1a1a2e",
      projection   = list(type = "mercator"),
      lonaxis      = list(range = c(-110, -10)),
      lataxis      = list(range = c(5, 60))
    ),
    paper_bgcolor = "#1a1a2e",
    font = list(color = "white")
  ) |>
  animation_opts(frame = 1000, transition = 0)

#.....save as a self-contained HTML — anyone can open this file.....
saveWidget(animated_storms,
           file          = here::here("week9", "outputs", "storms_animated.html"),
           selfcontained = TRUE)
```

::: {.callout-note}
`selfcontained = TRUE` bundles all JavaScript and CSS into a single ~3 MB file. Set it to `FALSE` if you are hosting the file on a web server — this produces a smaller HTML plus a `_files/` folder.
:::

### Embedding works automatically in Quarto

In a Quarto HTML document, you do **not** need `saveWidget()`. Just print the widget object inside a code chunk — Quarto handles embedding automatically when you click **Render**.

```{r embed-demo}
#| eval: true
#| echo: true
#| warning: false

# Assign and print — Quarto embeds this in the rendered HTML
trend_widget <- storms_clean |>
  mutate(decade = paste0(floor(year / 10) * 10, "s")) |>
  count(decade, status) |>
  plot_ly(
    x             = ~decade,
    y             = ~n,
    color         = ~status,
    colors        = RColorBrewer::brewer.pal(8, "Paired"),
    type          = "bar",
    text          = ~paste0("<b>", status, "</b><br>n = ", n),
    textposition  = "none",
    hoverinfo     = "text",
    hovertemplate = "%{text}<extra></extra>"
  ) |>
  layout(
    barmode = "stack",
    title   = "Atlantic Storm Counts by Decade and Status",
    xaxis   = list(title = "Decade"),
    yaxis = list(title = "", showticklabels = FALSE),
    legend  = list(title = list(text = "<b>Status</b>"))
  )

trend_widget   # printing here embeds it in the rendered output
```

