---
title: "Lecture 4.2 KEY"
subtitle: "Visualizing relationships"
author: "Your Name"
date: 2025-01-27
toc: true
---

::: {.callout-note}
This template follows [lecture 4.2 slides](../lecture-slides/lecture4.2-relationships-slides.qmd){target="_blank"}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

## Setup

```{r}
#| eval: false
#| echo: true
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(metajam) 
library(tidyverse)
library(ggExtra)
library(ggdensity)

#...................download data from DataOne...................
# you only need to do this once (then I recommend commenting it out)!
download_d1_data("https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fknb-lter-hbr%2F208%2F11%2F3b3cf7ea447cb875d7c7d68ebdfd24c7",
                 path = here::here("week4")) 

#  ~ NOTE: You should rename the downloaded folder to 'data/' so that it's ignored by .gitignore! ~

#....................read in downloaded files....................
stream_chem_all <- read_d1_files(here::here("week4", "data"))

#........................get the data file.......................
stream_chem_data <- stream_chem_all$data
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(metajam) 
library(tidyverse)

#...................download data from DataOne...................
# download_d1_data("https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fknb-lter-hbr%2F208%2F11%2F3b3cf7ea447cb875d7c7d68ebdfd24c7",
#                  path = here::here("course-materials", "lecture-slides", "data"))

#  ~ NOTE: You should rename the downloaded folder to 'data/' so that it's ignored by .gitignore! ~

#....................read in downloaded files....................
stream_chem_all <- read_d1_files(here::here("course-materials", "lecture-slides", "data", "https_pasta.lternet.edu_package_metadata_eml_knb-lter-hbr_208_11__HubbardBrook_weekly_stream_chemistry__csv"))

#........................get the data file.......................
stream_chem_data <- stream_chem_all$data
```

## Scatter plots

### Basic scatter plot

```{r}
#| eval: true
#| echo: true
#| warning: false
# basic scatter plot ----
p1 <- stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH)) + 
  geom_point(alpha = 0.5)

p1
```

### Alt 1: add a rug plot

```{r}
#| eval: true
#| echo: true
#| warning: false
# scatter + rug plot ----
p1 +
  geom_rug()
```

### Alt 2: marginal plot alternatives using `{ggExtra}`

```{r}
#| eval: true
#| echo: true
#| warning: false
# add marginal plots with {ggExtra} ----
# also try "density", "boxplot"
ggExtra::ggMarginal(p1, type = "histogram")
```

### Alt 3: scatter + marginal plot with 2+ groups

```{r}
#| eval: true
#| echo: true
#| warning: false
# scatter plot with points colored by site ----
p2 <- stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = site)) + 
  geom_point(alpha = 0.5) +
  theme(legend.position = "bottom")

# add marginal density plot ----
ggExtra::ggMarginal(p2, type = "density", groupFill = TRUE, groupColour = TRUE)
```

## Trend lines

### Default (<1000 data points): LOESS method

- similar to a moving average

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
# add a smoothed line ----
stream_chem_data |> 
  filter(waterYr %in% c(2011:2021)) |> 
  filter(site == "W5") |> 
  ggplot(aes(x = DOC, y = pH)) + 
  geom_point(alpha = 0.5) +
  geom_smooth()
```

### Line of best fit

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
# add a trend line ----
stream_chem_data |> 
  filter(waterYr %in% c(2011:2021)) |> 
  filter(site == "W5") |>  
  ggplot(aes(x = DOC, y = pH)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE)
```

## Visualizing a third numeric variable

### Bubble charts

- variation of a scatter plot for visualizing a third numeric variable
- be extra mindful:
    - primary focus = relationship between your x- and y-axis variables 
    - hard to compare the strengths of different associations (is there a better alternative graphic form?)
    - if the range of values mapped to size is small, your bubbles will look indistinguishable from one another
    - consider adjusting the size range of your scale
    
```{r}
#| eval: true
#| echo: true
#| warning: false
# bubble chart ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) + 
  geom_point(alpha = 0.5) +
  labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")
```

- adjust size range of bubbles

```{r}
#| eval: true
#| echo: true
#| warning: false
# scale size by area ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) + 
  geom_point(alpha = 0.5) +
  scale_size(range = c(1, 10)) +
  labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")
```

- IMPORTANT: don't scale size by radius (can be deceptive)

```{r}
#| eval: true
#| echo: true
#| warning: false
# don't scale size by radius ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) + 
  geom_point(alpha = 0.5) +
  scale_radius(range = c(1, 10)) +
  labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")
```

### Use color for third variable

```{r}
#| eval: true
#| echo: true
#| warning: false
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = Al_ICP)) + 
  geom_point(alpha = 0.5, size = 2) +
  scale_color_viridis_c() +
  labs(x = "DOC (mg/L)", color = "Al (mg/L)")
```

### But oftentimes best to just create two different plots

```{r}
#| eval: true
#| echo: true
#| warning: false
#| layout-ncol: 2
# effect of DOC on pH ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = site)) + 
  geom_point(alpha = 0.5)

# effect of DOC on Al_ICP ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = Al_ICP, y = pH, color = site)) + 
  geom_point(alpha = 0.5)
```

## Overplotting

- if you have too many points, scatter plots can be ineffective

```{r}
#| eval: true
#| echo: true
#| warning: false
# wayyyy too many points ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) + 
  geom_point()
```

### Initial strategies

- adjust size / transparency of points

```{r}
#| eval: true
#| echo: true
#| warning: false
# smaller + more transparent points ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) + 
  geom_point(size = 0.5, alpha = 0.3) 
```

- add a rug plot (or marginal plot)

```{r}
#| eval: true
#| echo: true
#| warning: false
# added rug plot ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) + 
  geom_point(size = 0.5, alpha = 0.3) +
  geom_rug()
```

- color by group

```{r}
#| eval: true
#| echo: true
#| warning: false
# added rug plot ----
ggplot(stream_chem_data, aes(x = SO4, y = pH, color = site)) + 
  geom_point(size = 0.5, alpha = 0.3) 
```

### Alt 1: 2d heatmaps

- e.g. like you're looking down on a histogram

```{r}
#| eval: true
#| echo: true
#| warning: false
# heatmap of 2d bin counts ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) +
  geom_bin2d() +
  scale_fill_viridis_c()
```

- or use hexagonal shapes
- also increase height of legend can make continuous scale easier to read

```{r}
#| eval: true
#| echo: true
#| warning: false
# heatmap of 2d bin counts ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) +
  geom_hex() +
  scale_fill_viridis_c() +
  guides(fill = guide_colorbar(title = "Count", 
                                barwidth = 1, barheight = 15))
```

### Alt 2: 2d density / contour plots

- e.g. like you're looking down on a density plot
- legend provides an estimate of the proportion of data points fall within a colored region (density of distribution of points sums to 1)
    - interpreting legend: 0-2% of of points fall within a 1x1 square in the darkest blue region, while 26-28% fall within a 1x1 square in the brightest yellow region

```{r}
#| eval: true
#| echo: true
#| warning: false
# contour plot ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) +
  geom_density_2d_filled()
```

### Alt 3: 2d density plots using `{ggdensity}`

- can be easier to interpret
- compute and plot the resulting highest density regions (HDRs)
- HDRs are computed to be the smallest such regions that bound that level of probability

```{r}
#| eval: true
#| echo: true
#| warning: false
#  ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) +
  ggdensity::geom_hdr() # update prob level e.g. `probs = c(0.99, 0.75, 0.5, 0.25)`
```


