---
title: "KEY - interactive data visualization"
subtitle: "plotly & htmlwidgets"
author: "Your Name"
date: today
toc: true
toc-depth: 3
editor_options: 
  chunk_output_type: inline
---

::: {.callout-note}
This template follows the [interactive data visualization](../lecture-slides/SLIDES-interactivity.qmd){target="_blank"} slides. Please be sure to cross-reference the slides, which contain important conceptual context and comparisons!
:::

## Setup

::: {.callout-note}
We use the `storms` dataset from `{dplyr}` throughout this lab — no external file needed. It contains Atlantic storm tracks with wind speed, pressure, storm status, and geographic position from 1975 to 2022.
:::

<!-- CODE RENDERS BUT DOES NOT RUN -->

```{r setup-echo}
#| eval: false
#| echo: true
#| warning: false
#| message: false

#..........................load packages.........................
library(tidyverse)
library(plotly)
library(htmlwidgets)
library(scales)

#.......................preview the dataset.......................
glimpse(storms)
```

<!-- CODE RUNS BUT DOES NOT RENDER (FOR FILE PATH) -->

```{r setup-eval}
#| eval: true
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(plotly)
library(htmlwidgets)
library(scales)
```

**Key variables in `storms`:**

| Variable | Type | Description |
|---|---|---|
| `name` | character | Storm name |
| `year`, `month`, `day`, `hour` | numeric | Date & time of observation |
| `lat`, `long` | numeric | Storm center position |
| `status` | factor | `"tropical depression"`, `"tropical storm"`, `"hurricane"`, etc. |
| `category` | numeric | Saffir-Simpson category (1–5; `NA` for non-hurricanes) |
| `wind` | numeric | Max sustained wind speed (knots) |
| `pressure` | numeric | Min central pressure (millibars) |
| `tropicalstorm_force_diameter` | numeric | Diameter (nautical miles) of TS-force winds |

```{r preview-data}
#| eval: true
#| echo: true
#| message: false

glimpse(storms)
```

## Wrangle data

We create a few focused subsets used throughout the lab.

```{r wrangle-data}
#| eval: true
#| echo: true
#| message: false

#........one observation per storm-year (peak intensity).........
# storms has multiple rows per storm (one per 6-hour observation)
# we summarise to peak wind per named storm per year
storms_peak <- storms |>
    filter(status %in% c("tropical depression", "tropical storm", "tropical wave")) |>
  group_by(name, year, status) |>
  summarise(
    peak_wind     = max(wind, na.rm = TRUE),
    min_pressure  = min(pressure, na.rm = TRUE),
    max_category  = max(category, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    # clean up category for labelling
    category_label = case_when(
      max_category == 1  ~ "Cat 1",
      max_category == 2  ~ "Cat 2",
      max_category == 3  ~ "Cat 3",
      max_category == 4  ~ "Cat 4",
      max_category == 5  ~ "Cat 5",
      TRUE               ~ status
    )
  )

#.............annual summaries by storm status...................
storms_annual <- storms |>
  filter(status %in% c("tropical depression", "tropical storm", "tropical wave")) |>
  group_by(year, status) |>
  summarise(
    mean_wind      = mean(wind, na.rm = TRUE),
    n_observations = n(),
    .groups = "drop"
  )

#...........filter to recent decades for cleaner plots...........
storms_2000 <- storms_peak |>
  filter(year >= 2000) |>
   mutate(
    status_label = recode(status,
      "tropical depression" = "Depression",
      "tropical storm"      = "Tropical Storm",
      "tropical wave"       = "Tropical Wave"
    ),
    status_label = factor(status_label,
      levels = c("Depression", "Tropical Storm", "Tropical Wave")),
    label = paste0(
      "<b>", name, " (", year, ")</b><br>",
      "Peak wind: ", peak_wind, " kt  ·  Min pressure: ", min_pressure, " mb"
    )
  )

pal <- c(
  "Depression"     = "#7FB3D3",
  "Tropical Storm" = "#E8936A",
  "Wave"           = "#82C99A"
)


glimpse(storms_peak)
```

---

## Part 1: `ggplotly()` — converting a static ggplot

See slides: [Two approaches](../lecture-slides/SLIDES-interactivity.qmd#two-approaches){target="_blank"} and [ggplotly()](../lecture-slides/SLIDES-interactivity.qmd#ggplotly-how){target="_blank"}

### Build a static ggplot

We start with a scatter plot of peak wind speed vs. minimum pressure for storms since 2000. 

```{r static-ggplot}
#| eval: true
#| echo: true
#| warning: false
#| fig-align: "center"
#| out-width: "80%"

#...................add formatted hover label....................
storms_2000 <- storms_2000 |>
  mutate(label = paste0(
    "<b>", name, " (", year, ")</b><br>",
    "Status: ", category_label, "<br>",
    "Peak wind: ", peak_wind, " kt<br>",
    "Min pressure: ", min_pressure, " mb"
  ))

#............................build plot..........................
p <- ggplot(storms_2000,
            aes(x = peak_wind, y = min_pressure,
                color = status_label, text = label)) +
  geom_jitter(size = 2.5, alpha = 0.65, width = 0.8, height = 0) +  # jitter x only
  scale_color_manual(values = pal, name = NULL) +
  labs(
    title = "Atlantic Storm Peak Intensity (2000–2022)",
    x     = "Peak Wind Speed (kt)",
    y     = "Minimum Pressure (mb)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey93"),
    legend.position  = "top",
    legend.text      = element_text(size = 11)
  )


p

```

### Convert to interactive with `ggplotly()`

```{r ggplotly-basic}
#| eval: true
#| echo: true
#| warning: false

ggplotly(p, tooltip = "text")
```

::: {.callout-tip}
**Try it:** Hover over any point to see the formatted label. Click a status category in the legend to hide it.Double click a status category to view only that category. Drag to zoom; double-click to reset. The toolbar in the top-right gives additional controls.
:::

### Customising `ggplotly()` output

A few small tweaks make `ggplotly()` output cleaner:

```{r ggplotly-polished}
#| eval: true
#| echo: true
#| warning: false

ggplotly(p, tooltip = "text") |>
  layout(
    legend     = list(orientation = "h", x = 0, y = 1.1),
    hoverlabel = list(bgcolor = "white", bordercolor = "grey80",
                      font = list(size = 12))
  ) |>
  config(displaylogo = FALSE,
         modeBarButtonsToRemove = list("select2d", "lasso2d"))
```

---

## Part 2: `plot_ly()` — native plotly for more control

See slides: [plot_ly() grammar](../lecture-slides/SLIDES-interactivity.qmd#plot-ly-grammar){target="_blank"} and [animation](../lecture-slides/SLIDES-interactivity.qmd#animation-intro){target="_blank"}

### 2A — Animated scatter: storm intensity across decades

The `frame` argument is the key to animation. Plotly creates one frame per unique value and adds a Play/Pause button and year slider automatically.

```{r animated-scatter}
#| eval: true
#| echo: true
#| warning: false


storms_peak <- storms |>
  group_by(name, year, status) |>
  summarise(
    peak_wind     = max(wind, na.rm = TRUE),
    min_pressure  = min(pressure, na.rm = TRUE),
    max_category  = max(category, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    # clean up category for labelling
    category_label = case_when(
      max_category == 1  ~ "Cat 1",
      max_category == 2  ~ "Cat 2",
      max_category == 3  ~ "Cat 3",
      max_category == 4  ~ "Cat 4",
      max_category == 5  ~ "Cat 5",
      TRUE               ~ status
    )
  )

storms_peak |>
  mutate(
    # build hover label
    hover_text = paste0(
      "<b>", name, "</b><br>",
      "Year: ", year, "<br>",
      "Status: ", category_label, "<br>",
      "Peak wind: ", peak_wind, " kt<br>",
      "Min pressure: ", min_pressure, " mb"
    )
  ) |>
  plot_ly(
    x         = ~peak_wind,
    y         = ~min_pressure,
    color     = ~status,
    frame     = ~year,           # one frame per year → animation
    text      = ~hover_text,
    hoverinfo = "text",
    type      = "scatter",
    mode      = "markers",
    marker    = list(size = 8, opacity = 0.7)
  ) |>
  layout(
    title  = "Atlantic Storm Peak Intensity by Year (1975–2022)",
    xaxis  = list(title = "Peak Wind Speed (knots)", range = c(0, 175)),
    yaxis  = list(title = "Min Central Pressure (mb)", range = c(880, 1020)),
    legend = list(title = list(text = "Storm status:"))
  ) |>
  animation_opts(
    frame   = 700,              # ms per year frame
    easing  = "linear",         # constant speed — clearest for data trends
    redraw  = FALSE
  ) |>
  animation_button(
    x = 1.08, xanchor = "right",
    y = 0,    yanchor = "bottom"
  )
```




###  2B: Bar chart: storm counts by status and decade

```{r bar-chart}
#| eval: true
#| echo: true
#| warning: false

storms_peak |>
  mutate(decade = paste0(floor(year / 10) * 10, "s")) |>
  count(decade, status) |>
  plot_ly(
    x         = ~decade,
    y         = ~n,
    color     = ~status,
    type      = "bar",
    text      = ~paste0(status, "<br>n = ", n),
    hoverinfo = "text"
  ) |>
  layout(
    barmode = "stack",
    title   = "Atlantic Storm Counts by Decade and Status",
    xaxis   = list(title = "Decade"),
    yaxis   = list(title = "Number of Named Storms"),
    legend  = list(title = list(text = "Storm status:"))
  )
```

---

## Part 3: Saving and embedding widgets

See slides: [saveWidget()](../lecture-slides/SLIDES-interactivity.qmd#htmlwidgets-save){target="_blank"} and [embedding in Quarto](../lecture-slides/SLIDES-interactivity.qmd#quarto-embed){target="_blank"}

### Save a widget to a standalone HTML file

```{r save-widget}
#| eval: false
#| echo: true

#...............assign the animated chart to an object...........
animated_storms <- storms_peak |>
  mutate(hover_text = paste0("<b>", name, "</b><br>",
                             "Year: ", year, "<br>",
                             "Peak wind: ", peak_wind, " kt")) |>
  plot_ly(x = ~peak_wind, y = ~min_pressure, color = ~status,
          frame = ~year, text = ~hover_text, hoverinfo = "text",
          type = "scatter", mode = "markers") |>
  layout(xaxis = list(title = "Peak Wind (knots)"),
         yaxis = list(title = "Min Pressure (mb)"),
         title = "Atlantic Storm Intensity (animated)")

#.....save as a self-contained HTML — anyone can open this file.....
saveWidget(animated_storms,
           file          = here::here("week9", "outputs", "storms_animated.html"),
           selfcontained = TRUE)
```

::: {.callout-note}
`selfcontained = TRUE` bundles all JavaScript and CSS into a single ~3 MB file. Set it to `FALSE` if you are hosting the file on a web server — this produces a smaller HTML plus a `_files/` folder.
:::

### Embedding works automatically in Quarto

In a Quarto HTML document, you do **not** need `saveWidget()`. Just print the widget object inside a code chunk — Quarto handles embedding automatically when you click **Render**.

```{r embed-demo}
#| eval: true
#| echo: true
#| warning: false

# Assign and print — Quarto embeds this in the rendered HTML
trend_widget <- storms_annual |>
  filter(status %in% c("tropical storm", "hurricane")) |>
  plot_ly(x = ~year, y = ~mean_wind, color = ~status,
          type = "scatter", mode = "lines+markers",
          text  = ~paste0(status, " (", year, "): ",
                          round(mean_wind, 1), " kt"),
          hoverinfo = "text") |>
  layout(title = "Mean Storm Wind Speed by Status",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Mean Wind (knots)"))

trend_widget   # printing here embeds it in the rendered output
```

