---
title: "Week 6 Lab: Exercise"
subtitle: "Scrollytelling"
description: "February 12^th^, 2026"
format: 
  closeread-html:
    code-copy: hover
    theme: 
      - closeread-styles.scss
---


::::{.cr-section layout="sidebar-right"}

@cr-scrollytelling 
  
:::{#cr-scrollytelling .body-text-xl style="margin-top: 200px; margin-left: 50px;"}
Have you ever wanted to go beyond the traditional structure of a qmd? Is the interactivity on your website not cutting it for you? Scrollytelling is a powerful data storytelling tool that allows you to invite your reader to "take a walk in the data's shoes". The `closeread` quarto extension provides flexibility on interactive customizations to tell your story the way you want. 
:::

:::{.body-text-l style="margin-left: 50px;"}
Let's walk through how you create your own scrollytelling story with `closeread`! 
:::

::::


::::{.cr-section layout="sidebar-right"}

@cr-setup

:::{#cr-setup .body-text-l style="margin-top: 200px; margin-left: 50px;"}
To start, make sure you have added the closeread quarto extension. Navigate to your project directory where you plan to use closeread. Scroll to find out what to paste into your terminal!
:::

:::{focus-on="cr-setup"}

:::{.body-text-m}
```{bash}
#| eval: false
quarto add qmd-lab/closeread
```
:::

<button onclick="navigator.clipboard.writeText('quarto add qmd-lab/closeread')" style="padding: 8px 16px; background: #047C90; color: white; border: none; border-radius: 4px; cursor: pointer; margin-top: 10px; font-family: 'Nunito', sans-serif;">Copy the above command here!</button>

:::{.body-text-m}  
<br>
Type **"Yes"** when prompted to allow the extension to execute code
:::

:::

::::


::::{.cr-section layout="sidebar-right"}

@cr-lets-get-started

:::{#cr-lets-get-started .body-text-xl .teal-text style="margin-top: 200px; margin-left: 75px;"}
Now that the setup is done, let's get started!!
:::

::::


::::{.cr-section layout="overlay-center"}

:::{focus-on="cr-title" style="height: 100vh;"}
:::

:::{#cr-title .body-text-xl .dark-blue-text style="padding: 40px;"}
**The Main Components**
:::

:::{focus-on="cr-stickies" style="height: 100vh;"}
:::

:::{#cr-stickies .body-text-l .dark-blue-text style="padding: 40px; text-align: left;"}
**The Main Components**

[**Stickies:**]{.teal-text} Think of the stickies as a sticky note - stuck in place! As the user starts to scroll, stickies remain in place. Stickies can take the form of text, an image, or a code chunk. 
:::

:::{focus-on="cr-triggers" style="height: 100vh;"}
:::

:::{#cr-triggers .body-text-l .dark-blue-text style="padding: 40px; text-align: left;"}
**The Main Components**

[**Stickies:**]{.teal-text} Think of the stickies as a sticky note - stuck in place! As the user starts to scroll, stickies remain in place. Stickies can take the form of text, an image, or a code chunk. 

[**Triggers:**]{.teal-text} Prompt the moving text, image, or plot that moves past your stickies.
:::

::::


::::{.cr-section layout="sidebar-right"}

[This is an example of a trigger! It moves as you scroll!]{.body-text-l} @cr-stickie-ex

[This is a second example of a trigger! Notice how our stickie is still there!]{.body-text-l} @cr-stickie-ex

:::{#cr-stickie-ex .body-text-xl .teal-text style="margin-left: 50px;"}
This is an example of a stickie!! As you scroll, the text will stay in place until you get to a new section. 
:::

::::


::::{.cr-section layout="sidebar-right"}

[Triggers can also be plots! Scroll to see]{.body-text-l} @cr-stickie-ex2

```{r}
#| message: false
#| warning: false
library(ggplot2)
library(palmerpenguins)

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Bill Length vs Bill Depth",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    color = "Species"
  ) +
  theme_minimal()
```

@cr-stickie-ex2

:::{#cr-stickie-ex2}
```{r}
#| message: false
#| warning: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Flipper Length vs Body Mass",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```
:::

::::


::::{.cr-section layout="sidebar-right"}

:::{focus-on="cr-section-syntax" .body-text-l style="margin-top: 200px; margin-left: 50px;"}
Every closeread document is built from sections. Within each section, you'll specify a layout that controls where your narrative and sticky content appear. There are five layout options to choose from.
:::

:::{#cr-section-syntax style="font-size: 28px; padding: 40px;"}
```
::::{.cr-section layout="LAYOUT-NAME"}
Content goes here!
::::
```
:::




:::{focus-on="cr-section-syntax"}
[**sidebar-left** (default)]{.teal-text}

Narrative scrolls on the left

[**sidebar-right** ]{.teal-text}

Narrative scrolls on the right

[**overlay-left**]{.teal-text}  
Full-screen sticky with floating narrative on left

[**overlay-center**]{.teal-text}  
Full-screen sticky with centered floating narrative

[**overlay-right**]{.teal-text}  
Full-screen sticky with floating narrative on right

<br><br>
[**Choose the layout that best showcases your story!**]{.dark-blue-text .body-text-m}
:::


::::


::::{.cr-section layout="overlay-center"}

:::{focus-on="cr-scroll-back" style="height: 50vh;"}
:::

:::{#cr-scroll-back .body-text-xl .dark-blue-text style="padding: 40px; text-align: center; margin-top: 40px"}
Scroll back up if the terms **stickie** or **trigger** aren't making sense yet. Otherwise, keep scrolling to find out how to code them!
:::

::::


::::{.cr-section layout="sidebar-left"}

:::{focus-on="cr-sticky-syntax"}
[To create a sticky, wrap your element in a fenced div and provide an ID prefixed with `cr-`. Stickies will look like this in your qmd file:]{.body-text-m}
:::

:::{#cr-sticky-syntax .body-text-m style="padding: 40px; margin-top:150px"}
[**Creating a Sticky:**]{.teal-text}

```
:::{#cr-YOUR-ID}
Your content goes here!
(text, image, plot, code, etc.)
:::
```

- Must start with `#cr-`
- ID must be unique
- Can contain any element
:::

:::{focus-on="cr-trigger-syntax"}
[To create a trigger, reference your sticky ID using the `@cr-` syntax. When this text scrolls into view, it will reveal your sticky!]{.body-text-m}
:::

:::{#cr-trigger-syntax .body-text-m style="padding: 40px;"}
[**Creating a Trigger:**]{.teal-text}

```
Your narrative text goes here. @cr-YOUR-ID
```

Or wrap multiple blocks:

```
:::{focus-on="cr-YOUR-ID"}
Your narrative text goes here.
More text here too!
:::
```

- `@cr-` for single paragraph triggers
- `focus-on=` for multi-block triggers
:::

::::


::::{.cr-section layout="sidebar-right"}

[Let's see it in action! This text you're reading right now is a **trigger**. It scrolls past while the sticky on the left stays fixed.]{.body-text-m} @cr-live-example

[See how this text is different from the sticky? This is a second trigger - it scrolls into view while the sticky stays put!]{.body-text-m} @cr-live-example

[And this is a third trigger! No matter how many triggers reference the same sticky, it stays locked in place until you scroll to a new section.]{.body-text-m} @cr-live-example

:::{#cr-live-example .body-text-m style="padding: 40px; border-left: 5px solid #047C90;"}
[**I am the STICKY!**]{.teal-text}

I stay fixed in place while the triggers on the right scroll past me.

In your `.qmd` file I look like this:

```
:::{#cr-live-example}
I am the sticky!
:::
```

And my triggers look like this:

```
This text is a trigger! @cr-live-example
```
:::

::::
