---
title: "Interactive data viz, Alternative chart type practice"
subtitle: "Week 9"
description: "Wednesday March 4^th^, 2026"
---

```{r interactive-viz-ex}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| out-width: "70%"
#| out-height: "70%"
#..........................load packages.........................
library(tidyverse)
library(plotly)
library(htmlwidgets)
library(scales)

#........one observation per storm-year (peak intensity).........
# storms has multiple rows per storm (one per 6-hour observation)
# we summarise to peak wind per named storm per year
storms_peak <- storms |>
    filter(status %in% c("tropical depression", "tropical storm", "tropical wave")) |>
  group_by(name, year, status) |>
  summarise(
    peak_wind     = max(wind, na.rm = TRUE),
    min_pressure  = min(pressure, na.rm = TRUE),
    max_category  = max(category, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    # clean up category for labelling
    category_label = case_when(
      max_category == 1  ~ "Cat 1",
      max_category == 2  ~ "Cat 2",
      max_category == 3  ~ "Cat 3",
      max_category == 4  ~ "Cat 4",
      max_category == 5  ~ "Cat 5",
      TRUE               ~ status
    )
  )

#...........filter to recent decades for cleaner plots...........
storms_2000 <- storms_peak |>
  filter(year >= 2000) |>
   mutate(
    status_label = recode(status,
      "tropical depression" = "Depression",
      "tropical storm"      = "Tropical Storm",
      "tropical wave"       = "Tropical Wave"
    ),
    status_label = factor(status_label,
      levels = c("Depression", "Tropical Storm", "Tropical Wave")),
    label = paste0(
      "<b>", name, " (", year, ")</b><br>",
      "Peak wind: ", peak_wind, " kt  ·  Min pressure: ", min_pressure, " mb"
    )
  )

pal <- c(
  "Depression"     = "#7FB3D3",
  "Tropical Storm" = "#E8936A",
  "Wave"           = "#82C99A"
)

#...................add formatted hover label....................
storms_2000 <- storms_2000 |>
  mutate(label = paste0(
    "<b>", name, " (", year, ")</b><br>",
    "Status: ", category_label, "<br>",
    "Peak wind: ", peak_wind, " kt<br>",
    "Min pressure: ", min_pressure, " mb"
  ))

#............................build plot..........................
p <- ggplot(storms_2000,
            aes(x = peak_wind, y = min_pressure,
                color = status_label, text = label)) +
  geom_jitter(size = 2.5, alpha = 0.65, width = 0.8, height = 0) +  # jitter x only
  scale_color_manual(values = pal, name = NULL) +
  labs(
    title = "Atlantic Storm Peak Intensity (2000–2022)",
    x     = "Peak Wind Speed (kt)",
    y     = "Minimum Pressure (mb)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey93"),
    legend.position  = "top",
    legend.text      = element_text(size = 11)
  )


ggplotly(p, tooltip = "text") |> 
  config(displayModeBar = FALSE)
```

<br>

## {{< fa clipboard-list title="a clipboard with a To Do list" >}} Pre-class Prep {#pre-class-prep}

### A. Install required packages

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: false
library(plotly) # for making interactive graphs
library(htmlwidgets) # for using JavaScript data viz libraries in R
library(waffle) # for making waffle (square pie) charts in R
library(ggbump) # a geom for creating bump charts
```

### B. Download .qmd templates 

Download all necessary templates (links in the table below) and save them to a `EDS-240-data-viz-examples/week9/` folder. 

## {{< fa person-chalkboard title="a person pointing at a chalkboard" >}} Lecture Materials {#lecture-materials}

| {{< fa person-chalkboard title="A person motioning at a chalkboard" >}} Lecture slides   | {{< fa file-code title="A code file" >}} Code-along template | {{< fa key title="A key" >}} Code-along key |
|------------------------------------------------------------------------------------------|--------------------------------------------------------------|---------------------------------------------------|
| [Lecture 9.1: interactive data viz](lecture-slides/SLIDES-interactivity.qmd){target="_blank"} | *coming soon* | [9.1 key](keys/KEY-interactivity.qmd){target="_blank"} |
| [Lecture 9.2: alternative chart types](lecture-slides/SLIDES-alt-charts.qmd){target="_blank"} | *coming soon* | *coming soon* |
: {.hover .bordered tbl-colwidths="[50,25,25]"} 

## {{< fa clipboard-list title="a clipboard with a To Do list" >}} Pre-lab Prep {#pre-lab-prep} 
Similar to last week's lab, we'll be discussing progress on our data visualizations and providing constructive feedback to one another.  

### A. Prepare your data viz slide 

Find your assigned slide from the correct Google Slide deck (see table, under the [Lab Materials](week9.qmd#lab-materials) section, below) and add the following:

1. ***(the final) one*** of your three visualizations (i.e. don't choose the same viz that you brought to the last two labs section)
2. the question that you're visualization helps to answer

## {{< fa comments title="two speech bubbles" >}} Lab Materials {#lab-materials}


| {{< fa person-chalkboard title="A person motioning at a chalkboard" >}} Lab Slides | {{< fa file-code title="A code file" >}} Exercise instructions | {{< fa key title="A key" >}} Exercise solutions |
|-------------------|-------------------|--------------|
| [Lab 9 slides: peer feedback](lab/LAB-SLIDES-peer-feedback3.qmd){target="_blank"} | [Google Slides](https://docs.google.com/presentation/d/1cmCNixyCyl_nPlyaolT1xpIoWLlJwOOBbJqsG_XvnGI/edit?usp=sharing){target="_blank"} | [Peer Feedback Form](https://docs.google.com/document/d/1xV-YF5PyfoZA_HXmiHC2NqaS5t5tTb8C_AtndsoM0JA/copy)  |
: {.hover .bordered tbl-colwidths="[50,25,25]"} 

## {{< fa pencil title="pencil" >}} Assignment Reminders {#assignment-reminders}

| Assignment Type | Assignment Title                                    | Date Assigned  | Date Due                   |
|-----------------|-----------------------------------------------------|----------------|----------------------------|
| FPM             | [Final Project Milestone #4](assignments/FPM4.qmd)  | Wed 02/25/2026 | Sun 03/08/2026, 11:59pm PT |
: {.hover .bordered tbl-colwidths="[20,30,20,30]"}

