---
title: "Interactive data viz, Alternative chart type practice"
subtitle: "Week 9"
description: "Wednesday March 4^th^, 2026"
---

```{r interactive-viz-ex}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| out-width: "70%"
#| out-height: "70%"
#..........................load packages.........................
library(tidyverse)
library(plotly)
library(htmlwidgets)
library(scales)
library(dplyr)

storms_clean <- storms %>%
  mutate(status = tools::toTitleCase(gsub("-", " ", status))) %>%
  filter(!is.na(wind), !is.na(pressure), !is.na(status)) %>%
  filter(wind > 0, pressure > 800, pressure < 1030) %>% 
    group_by(name, year, status) %>%
  summarise(
    wind     = mean(wind, na.rm = TRUE),
    pressure = mean(pressure, na.rm = TRUE),
    .groups  = "drop"
  )

ggplot_static <- storms_clean %>%
  ggplot(aes(
    x     = pressure,
    y     = wind,
    color = status,
    text  = paste0(
      "<b>", name, " (", year, ")</b><br>",
      "Status: ", status, "<br>",
      "Wind: ", round(wind,1), " kt<br>",
      "Pressure: ", round(pressure,1), " mb"
    )
  )) +
  geom_jitter(alpha = 0.35, size = 0.9, width = 1.5, height = 1.5) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title  = "Atlantic Storm Intensity by Status",
    x      = "Pressure (mb)",
    y      = "Wind Speed (knots)",
    color  = "Status"
  ) +
  theme_minimal(base_size = 13)

ggplotly(ggplot_static, tooltip = "text") |>
  layout(
    hoverlabel = list(bgcolor = "white", bordercolor = "grey80",
                      font = list(size = 12))
  ) |>
  config(displaylogo = FALSE,
         modeBarButtonsToRemove = list("select2d", "lasso2d"))

```

<br>

## {{< fa clipboard-list title="a clipboard with a To Do list" >}} Pre-class Prep {#pre-class-prep}

### A. Install required packages

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: false
library(plotly) # for making interactive graphs
library(htmlwidgets) # for using JavaScript data viz libraries in R
library(waffle) # for making waffle (square pie) charts in R
library(ggbump) # a geom for creating bump charts
```

### B. Download .qmd templates 

Download all necessary templates (links in the table below) and save them to a `EDS-240-data-viz-examples/week9/` folder. 

## {{< fa person-chalkboard title="a person pointing at a chalkboard" >}} Lecture Materials {#lecture-materials}

| {{< fa person-chalkboard title="A person motioning at a chalkboard" >}} Lecture slides   | {{< fa file-code title="A code file" >}} Code-along template | {{< fa key title="A key" >}} Code-along key |
|------------------------------------------------------------------------------------------|--------------------------------------------------------------|---------------------------------------------------|
| [Lecture 9.1: interactive data viz](lecture-slides/SLIDES-interactivity.qmd){target="_blank"} | [9.1 template](https://github.com/EDS-240-Data-Viz/code-along-templates/blob/main/TEMPLATE-interactivity.qmd){target="_blank"} | [9.1 key](keys/KEY-interactivity.qmd){target="_blank"} |
| [Lecture 9.2: alternative chart types](lecture-slides/SLIDES-alt-charts.qmd){target="_blank"} | [9.2 template](https://github.com/EDS-240-Data-Viz/code-along-templates/blob/main/TEMPLATE-alt-charts.qmd){target="_blank"} | [9.2 key](keys/KEY-alt-charts.qmd){target="_blank"} |
: {.hover .bordered tbl-colwidths="[50,25,25]"} 

## {{< fa clipboard-list title="a clipboard with a To Do list" >}} Pre-lab Prep {#pre-lab-prep} 
Similar to last week's lab, we'll be discussing progress on our data visualizations and providing constructive feedback to one another.  

### A. Prepare your data viz slide 

Find your assigned slide from the correct Google Slide deck (see table, under the [Lab Materials](week9.qmd#lab-materials) section, below) and add the following:

1. ***(the final) one*** of your three visualizations (i.e. don't choose the same viz that you brought to the last two labs section)
2. the question that you're visualization helps to answer

## {{< fa comments title="two speech bubbles" >}} Lab Materials {#lab-materials}


| {{< fa person-chalkboard title="A person motioning at a chalkboard" >}} Lab Slides | {{< fa file-code title="A code file" >}} Exercise instructions | {{< fa key title="A key" >}} Exercise solutions |
|-------------------|-------------------|--------------|
| [Lab 9 slides: peer feedback](lab/LAB-SLIDES-peer-feedback3.qmd){target="_blank"} | [Google Slides](https://docs.google.com/presentation/d/1cmCNixyCyl_nPlyaolT1xpIoWLlJwOOBbJqsG_XvnGI/edit?usp=sharing){target="_blank"} | [Peer Feedback Form](https://docs.google.com/document/d/1xV-YF5PyfoZA_HXmiHC2NqaS5t5tTb8C_AtndsoM0JA/copy)  |
: {.hover .bordered tbl-colwidths="[50,25,25]"} 

## {{< fa pencil title="pencil" >}} Assignment Reminders {#assignment-reminders}

| Assignment Type | Assignment Title                                    | Date Assigned  | Date Due                   |
|-----------------|-----------------------------------------------------|----------------|----------------------------|
| FPM             | [Final Project Milestone #4](assignments/FPM4.qmd)  | Wed 02/25/2026 | Sun 03/08/2026, 11:59pm PT |
: {.hover .bordered tbl-colwidths="[20,30,20,30]"}

