---
title: "Week 1 Discussion: Exercise"
subtitle: "Data wrangling with the `{tidyverse}`"
description: "Tuesday January 7^th^, 2025"
---

::: {.callout-note}
It's up to you to organize your own `week1-lab.qmd` file (i.e. there is no template). You may (should) discuss and work through today's exercise with a partner (or two!).
:::

[**Your goal is to transform the raw fracking data into a wrangled data frame that looks like this (only the first 6 rows are shown here):**]{.teal-text}

```{r}
#| eval: true
#| echo: false
#| warning: false
#| message: false

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(janitor)
library(usdata)

#......................import fracking data......................
fracking <- read_csv(here::here("course-materials", "lab", "week1", "data", "registryupload_1.csv"))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                        clean/wrangle fracking data                       ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

fracking_clean <- fracking |> 
  
  # clean column names ----
  janitor::clean_names() |> 
  
  # clean up dates ----
  mutate(job_start_date = str_remove(job_start_date, " AM")) |> # remove 'AM' from string
  mutate(datetime_start = mdy_hms(job_start_date)) |> # convert from string to date (save to new col)
  mutate(year = year(datetime_start)) |> # create 'year' col from date

  # select relevant cols ----
  select(datetime_start, year, state_name, well_name, total_base_water_volume) |> 
  
  # filter out non-state names ----
  filter(!state_name %in% c("Beaver", "Beckham", "Harper", "Hemphill", "Midland", "Red River", "Roosevelt", "Rusk", "State", "WARD")) |> 
  
  # make all words title case ----
  mutate(state_name = str_to_title(state_name)) |> 
  
  # fix misspelled state names ----
  mutate(state_name = case_when(
    state_name == "Colordao" ~ "Colorado",
    state_name == "Loiusiana" ~ "Louisiana",
    state_name == "Louisianna" ~ "Louisiana",
    state_name == "Lousiana" ~ "Louisiana",
    state_name == "New Mexcio" ~ "New Mexico",
    state_name == "Norh Dakota" ~ "North Dakota",
    state_name == "Norht Dakota" ~ "North Dakota",
    state_name == "North  Dakota" ~ "North Dakota",
    state_name == "North Dakata" ~ "North Dakota",
    state_name == "North Dakotta" ~ "North Dakota",
    state_name == "Noth Dakota" ~ "North Dakota",
    state_name == "Pennslvania" ~ "Pennsylvania",
    state_name == "Pennsylavania" ~ "Pennsylvania",
    state_name == "Pennsylvanya" ~ "Pennsylvania",
    state_name == "Penssylvania" ~ "Pennsylvania",
    state_name == "Texasa" ~ "Texas",
    state_name == "Texs" ~ "Texas", 
    state_name == "West Viginia" ~ "West Virginia",
    state_name == "Wyominng" ~ "Wyoming", 
    TRUE ~ state_name # copy over rest of state names from as-is
  )) |> 
  
  # remove rows that have a '?' mark ----
  filter(!str_detect(string = state_name, pattern = "\\?")) |> # `?` is a special chr; escape with `\\` prefix
  
  # mutate abbreviations to full state names ----
  mutate(state_name = ifelse(test = str_length(state_name) == 2, # if string in 'sn' col is 2 chrs long
                     yes = usdata::abbr2state(state_name), # replace abbreviation with full state name 
                     no = state_name)) |> # if string in 'sn' col is not 2 chrs long, keep state name as-is
  
  # create a column of just state abbreviations ----
  mutate(state_abb = usdata::state2abbr(state_name)) |> 
  
  # move 'state_abb' col after state_name col ----
  relocate(state_abb, .after = state_name) |> 
  
  # rename 'total_base_water_volume' to 'total_base_water_volume_gal' for clarity ----
  rename(total_base_water_volume_gal = total_base_water_volume) |> 

  # remove obs that don't have a measurement for 'total_base_water_volume' (NA) ----
  drop_na(total_base_water_volume_gal) |> 
  
  # reorder rows from earliest to most recent `datetime_start` value ----
  arrange(datetime_start) |> 
  
  head()

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                           convert to kable table                         ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

DT::datatable(fracking_clean)
```


<br>

1. [**Load the `{tidyverse}`, `{janitor}`, and `{usdata}` packages, read in `registryupload_1.csv`, and perform some basic data exploration.**]{.teal-text} Use the `{here}` package to read in your data and check out things like the dimensions, structure, types of variables, unique observations, etc. Read through the [metadata](https://drive.google.com/file/d/1n-eiTKPzMo8di4cjG1A94vhkMLMqFY62/view?usp=sharing){target="_blank"}, which accompanied our `registryupload_1.csv` file when downloaded from FracFocus.

2. [**Consider what data *cleaning* you might need to perform.**]{.teal-text} Discuss any curious data formatting (e.g. data types, inconsistent observation names, missing values, etc.) that you discovered during your data exploration step, above. What might you need to address in your next data wrangling steps?

::: {.callout-important}
## Pause for class discussion
Take a moment to share some of your findings with the rest of the class.
:::

3. [**Clean / wrangle the data!**]{.teal-text} Using the packages loaded in step 1, perform the following operations (***NOTE:** Each step below corresponds to a single line of code*):
    a. convert all column names from CamelCase to snake_case (for ease of readability and to adhere to the [Tidyverse style guide](https://style.tidyverse.org/syntax.html#object-names){target="_blank"} recommendation)
    b. remove `"AM"` from the `job_start_date` column observations
    c. convert observations in the `job_start_date` column from character strings to datetime objects and save them to a new column named `datetime_start`
    d. add a column named `year` that contains only the year from `datetime_start`
    e. keep only the necessary columns (`datetime_start`, `year`, `state_name`, `well_name`, `total_base_water_volume`)
    f. convert all observations in the `state_name` column to Title Case
    g. remove any rows where the state name ends in `?` 
    h. convert any state abbreviations to full names in the `state_name` column (**Hint:** use a combination of the `mutate()`, `ifelse()` & `usdata::abbr2state()` functions)
    i. correct any misspelled state names in the `state_name` column (**Note:** figure out how to update the names of a few misspelled states, the feel free to copy the complete code for this step from the expandable section, below)
    j. remove any rows that do not have a true US state name in the `state_name` column
    k. add a column named `state_abb` with just US state abbreviations
    l. move the `state_abb` column so that it sits immediately after the `state_name` column
    m. rename `total_base_water_volume` as `total_base_water_volume_gal` for clarity
    n. remove any observations that don't have a measurement for `total_base_water_volume_gal` (i.e. NA)
    o. reorder rows from earliest to most recent `datetime_start` values
  
::: {.callout-tip collapse=true}
## Expand for code to fix misspelled state names
```{r}
#| eval: false
#| echo: true
mutate(state_name = case_when(
    state_name == "Colordao" ~ "Colorado",
    state_name == "Loiusiana" ~ "Louisiana",
    state_name == "Louisianna" ~ "Louisiana",
    state_name == "Lousiana" ~ "Louisiana",
    state_name == "New Mexcio" ~ "New Mexico",
    state_name == "Norh Dakota" ~ "North Dakota",
    state_name == "Norht Dakota" ~ "North Dakota",
    state_name == "North  Dakota" ~ "North Dakota",
    state_name == "North Dakata" ~ "North Dakota",
    state_name == "North Dakotta" ~ "North Dakota",
    state_name == "Noth Dakota" ~ "North Dakota",
    state_name == "Pennslvania" ~ "Pennsylvania",
    state_name == "Pennsylavania" ~ "Pennsylvania",
    state_name == "Pennsylvanya" ~ "Pennsylvania",
    state_name == "Penssylvania" ~ "Pennsylvania",
    state_name == "Texasa" ~ "Texas",
    state_name == "Texs" ~ "Texas", 
    state_name == "West Viginia" ~ "West Virginia",
    state_name == "Wyominng" ~ "Wyoming", 
    TRUE ~ state_name # copy over rest of state names from as-is
  ))
```
:::



